Faithful Variable Screening for High-Dimensional Convex Regression

### 2014-07-22 tex/scam/additive_faithful.tex
minor edits

% OLD lemma, requires product distribution
% \begin{lemma}
% \label{lem:general_int_reduction}
% Let $F$ be a product distribution on $\mathbf{C}=[0,1]^s$ with density function $p$ which is positive on $\mathbf{C}$. Let
% $X=(X_1,...,X_s) \sim F$. Let $f: \mathbf{C} \rightarrow \R$ be an
% integrable function.
% Let 
% \begin{equation}
% f_k^*, \mu^* \coloneqq \argmin_{f_1,\ldots, f_s, \mu} \Bigl\{ \E \bigl( f(X)  -
%  \sum_{k=1}^s f_k(X_k) -\mu \bigr)^2 
%         \,:\, \E f_k(X_k) = 0\Bigr\}.
% \end{equation}
% Then $\mu^* = \E f(X)$ and $f^*_k(x_k) = \E[ f(X) \given x_k] - \E f(X)$ and this solution is unique.
% \end{lemma}

\begin{lemma}
\label{lem:general_int_reduction}
Let $F$ be a distribution on $C=[0,1]^p$ with a positive density
function $p$. Let $f: C \rightarrow \R$ be an integrable function,
and define 
\begin{equation}
f^*_1,...,f^*_s, \mu^* \coloneqq 
\arg\min \left\{ \E \Bigl( f(X) - \mu - \sum_{k=1}^p f_k(X_k)\Bigr)^2 \,:\,
\E f_k(X_k) = 0,\; \forall k=1,\ldots, p \right\}.
\end{equation}
Then $\mu^* = \E f(X)$,
\begin{equation}
\label{eq:backfit}
f^*_k(x_k) = \E\Bigl[ f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) \given
x_k\Bigr] - \E f(X) 
\end{equation}
and this solution is unique.
\end{lemma}

@@

\begin{proof}
  Let $f^*_1,...,f^*_s, \mu^*$ be the minimizers as defined.  We first
  show that the optimal $\mu$ is $\mu^* = \E f(X)$ for any $f_1, ...,
  f_k$ such that $\E f_k(X_k) = 0$. This follows from the stationarity
  condition, which states that $\mu^* = \E[ f(X) - \sum_k f_k(X_k)] =
  \E[ f(X) ]$. Uniqueness is apparent because the second derivative is
  strictly larger than zero and strong convexity is guaranteed.

@@

Suppose we have the additive approximation
\begin{equation}
\label{eqn:unconstrained_additive}
f_k^*, \mu^* \coloneqq \argmin_{f_1,\ldots, f_s, \mu} \Bigl\{ 
             \E \Bigl[\Bigl( f(X) - \sum_{k=1}^p f_k(X_k) -\mu \Bigr)^2 \Bigr]
         \,:\, \E f_k(X_k) = 0 \Bigr\}.
\end{equation}
We say that $f$ is \textit{additively faithful} under $F$ in case
$f^*_k = 0$ implies that $f$ does not depend on coordinate $k$.
\end{definition}

@@

Let us use the shorthand notation that $r(\mathbf{x}_{-k}) = \sum_{k'
  \neq k} f_{k'}(x_{k'})$ and assume without loss of generality that
$\mu = 0$. We then assume that for all $x_k$,
\begin{equation}
 \E[ f(X) - r(X_{-k})  \given x_k] \equiv 
 \int_{\mathbf{x}_{-k}}  p(\mathbf{x}_{-k} \given x_k ) 
 \big(f(\mathbf{x}) - r(\mathbf{x}_{-k}) \big) = 0.
\end{equation}
We let $p'(\mathbf{x}_{-k} \given x_k)$ denote 
$\frac{\partial p(\mathbf{x}_{-k} \given x_k)}{\partial x_k}$ and 
$p''(\mathbf{x}_{-k} \given x_k)$ denote 
$\frac{\partial^2 p(\mathbf{x}_{-k} \given x_k)}{\partial x_k^2}$ and
likewise for $f'(x_k, \mathbf{x}_{-k})$ and $f''(x_k,
\mathbf{x}_{-k})$. We then differentiate under the integral, valid
because all functions are bounded, and obtain
\begin{gather}
\int_{\mathbf{x}_{-k}} p'(\mathbf{x}_{-k} \given x_k) 
\big( f(\mathbf{x}) - r(\mathbf{x}_{-k}) \big) + 
p(\mathbf{x}_{-k} \given x_k) f'(x_k, \mathbf{x}_{-k}) d \mathbf{x}_{-k}  = 0 
\label{eqn:integral1a} \\
\int_{\mathbf{x}_{-k}} p''(\mathbf{x}_{-k} \given x_k) 
\big( f(\mathbf{x}) - r(\mathbf{x}_{-k}) \big)  + 2 p'(\mathbf{x}_{-k} \given x_k) f'(x_k, \mathbf{x}_{-k}) +
p(\mathbf{x}_{-k} \given x_k) f''(x_k, \mathbf{x}_{-k}) d\mathbf{x}_{-k}  = 0 .
\end{gather}

@@

This theorem naturally suggests a two-stage screening procedure for
variable selection.  In the first stage we fit a sparse convex additive model.
In the second stage, for every variable marked as irrelevant in the first stage,
we fit a univariate \emph{concave} function on the residual for
that variable. We refer to this procedure as AC/DC (additive
convex / decoupled concave) and describe it in more detail in
Section~\ref{sec:acdc}.

@@

\begin{proof}[Proof of theorem~\ref{thm:acdc_faithful}]
Fix $k$. Let $f^*_k$ and $g^*_k$'s be defined as in
equation~\ref{eqn:convex_additive} and
equation~\ref{eqn:concave_postprocess}.  Then
we have that
\begin{align*}
f^*_k &= \arg\min_{f_k} \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - f_k \big)^2 
    \,:\, f_k \in  \mathcal{C}^1,\, \E f_k(X_k) = 0 
  \Big\} \\
g^*_k &= \arg\min_{g_k} \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \big)^2 
    \,:\, g_k \in \mh \mathcal{C}^1,\, \E g_k(X_k) = 0 
  \Big\}
\end{align*}
Therefore, $f^*_k = 0$ and $g^*_k = 0$ implies that $\E[ f(X) -
\sum_{k'\neq k} f^*_{k'}(X_{k'}) \given x_k] = 0$.  Now we use
Lemma~\ref{cor:faithfulness_extension} with $\phi(\mathbf{x}_{-k}) =
f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'})$ and conclude that
$f^*_k = 0$ and $g^*_k = 0$ together imply that $f$ does not depend on $x_k$.
\end{proof}

>>>>

% OLD lemma, requires product distribution
% \begin{lemma}
% \label{lem:general_int_reduction}
% Let $F$ be a product distribution on $\mathbf{C}=[0,1]^s$ with density function $p$ which is positive on $\mathbf{C}$. Let
% $X=(X_1,...,X_p) \sim F$. Let $f: \mathbf{C} \rightarrow \R$ be an
% integrable function.
% Let 
% \begin{equation}
% f_k^*, \mu^* \coloneqq \argmin_{f_1,\ldots, f_p, \mu} \Bigl\{ \E \bigl( f(X)  -
%  \sum_{k=1}^s f_k(X_k) -\mu \bigr)^2 
%         \,:\, \E f_k(X_k) = 0\Bigr\}.
% \end{equation}
% Then $\mu^* = \E f(X)$ and $f^*_k(x_k) = \E[ f(X) \given x_k] - \E f(X)$ and this solution is unique.
% \end{lemma}

\begin{lemma}
\label{lem:general_int_reduction}
Let $F$ be a distribution on $C=[0,1]^p$ with a positive density
function $p$. Let $f: C \rightarrow \R$ be an integrable function,
and define 
\begin{equation}
f^*_1,...,f^*_p, \mu^* \coloneqq 
\arg\min \left\{ \E \Bigl( f(X) - \mu - \sum_{k=1}^p f_k(X_k)\Bigr)^2 \,:\,
\E f_k(X_k) = 0,\; \forall k=1,\ldots, p \right\}.
\end{equation}
Then $\mu^* = \E f(X)$,
\begin{equation}
\label{eq:backfit}
f^*_k(x_k) = \E\Bigl[ f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) \given
x_k\Bigr] - \E f(X) 
\end{equation}
and this solution is unique.
\end{lemma}

@@

\begin{proof}
  Let $f^*_1,...,f^*_p, \mu^*$ be the minimizers as defined.  We first
  show that the optimal $\mu$ is $\mu^* = \E f(X)$ for any $f_1, ...,
  f_k$ such that $\E f_k(X_k) = 0$. This follows from the stationarity
  condition, which states that $\mu^* = \E[ f(X) - \sum_k f_k(X_k)] =
  \E[ f(X) ]$. Uniqueness is apparent because the second derivative is
  strictly larger than zero and strong convexity is guaranteed.

@@

Suppose we have the additive approximation
\begin{equation}
\label{eqn:unconstrained_additive}
f_k^*, \mu^* \coloneqq \argmin_{f_1,\ldots, f_p, \mu} \Bigl\{ 
             \E \Bigl[\Bigl( f(X) - \sum_{k=1}^p f_k(X_k) -\mu \Bigr)^2 \Bigr]
         \,:\, \E f_k(X_k) = 0 \Bigr\}.
\end{equation}
We say that $f$ is \textit{additively faithful} under $F$ in case
$f^*_k = 0$ implies that $f$ does not depend on coordinate $k$.
\end{definition}

@@

Let us use the shorthand notation that $r(\mathbf{x}_{-k}) = \sum_{k'
  \neq k} f_{k'}(x_{k'})$ and assume without loss of generality that
$\mu^* = E[f(X)] = 0$. We then assume that for all $x_k$,
\begin{equation}
 \E[ f(X) - r(X_{-k})  \given x_k] \equiv 
 \int_{\mathbf{x}_{-k}}  p(\mathbf{x}_{-k} \given x_k ) 
 \big(f(\mathbf{x}) - r(\mathbf{x}_{-k}) \big) = 0.
\end{equation}
We let $p'(\mathbf{x}_{-k} \given x_k)$ denote 
$\frac{\partial p(\mathbf{x}_{-k} \given x_k)}{\partial x_k}$ and 
$p''(\mathbf{x}_{-k} \given x_k)$ denote 
$\frac{\partial^2 p(\mathbf{x}_{-k} \given x_k)}{\partial x_k^2}$ and
likewise for $f'(x_k, \mathbf{x}_{-k})$ and $f''(x_k,
\mathbf{x}_{-k})$. We then differentiate under the integral, valid
because all functions are bounded, and obtain
\begin{gather}
\int_{\mathbf{x}_{-k}} p'(\mathbf{x}_{-k} \given x_k) 
\big( f(\mathbf{x}) - r(\mathbf{x}_{-k}) \big) + 
p(\mathbf{x}_{-k} \given x_k) f'(x_k, \mathbf{x}_{-k}) d \mathbf{x}_{-k}  = 0 
\label{eqn:integral1a} \\
\int_{\mathbf{x}_{-k}} p''(\mathbf{x}_{-k} \given x_k) 
\big( f(\mathbf{x}) - r(\mathbf{x}_{-k}) \big)  + 2 p'(\mathbf{x}_{-k} \given x_k) f'(x_k, \mathbf{x}_{-k}) +
p(\mathbf{x}_{-k} \given x_k) f''(x_k, \mathbf{x}_{-k}) d\mathbf{x}_{-k}  = 0 .
\end{gather}

@@

@@

\begin{proof}[Proof of theorem~\ref{thm:acdc_faithful}]
Fix $k$. Let $f^*_k$ and $g^*_k$ be defined as in
equation~\ref{eqn:convex_additive} and
equation~\ref{eqn:concave_postprocess}.  Then
we have that
\begin{align*}
f^*_k &= \arg\min_{f_k} \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - f_k \big)^2 
    \,:\, f_k \in  \mathcal{C}^1,\, \E f_k(X_k) = 0 
  \Big\} \\
g^*_k &= \arg\min_{g_k} \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \big)^2 
    \,:\, g_k \in \mh \mathcal{C}^1,\, \E g_k(X_k) = 0 
  \Big\}
\end{align*}
Therefore, $f^*_k = 0$ and $g^*_k = 0$ implies that $\E[ f(X) -
\sum_{k'\neq k} f^*_{k'}(X_{k'}) \given x_k] = 0$.  Now we use
Lemma~\ref{cor:faithfulness_extension} with $\phi(\mathbf{x}_{-k}) =
f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'})$ and conclude that
$f^*_k = 0$ and $g^*_k = 0$ together imply that $f$ does not depend on $x_k$.
\end{proof}
